---
title: "The NWM Reanalysis API"
vignette: >
  %\VignetteIndexEntry{pkgdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  maessage = FALSE,
  fig.width  = 6,
  fig.height = 4#,
  #eval=nzchar(Sys.getenv("BUILD_VIGNETTES"))
)

oldoption <- options(scipen = 9999)
options(scipen = 9999)
```

```{r setup, echo = FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(lubridate)
library(viridis)
library(nwmHistoric)

plot_here = function(flows, daily = TRUE){ 
  if(daily){
    flows = aggregate_ymd(flows)
  } else{
    flows$ymd = flows$time
  }
  
  ggplot(data = flows, aes(x = ymd, y = flow)) +
  geom_line(size = .25) + 
  theme_minimal() +
  labs(y = "Qsim (cms)",
       x = paste0("DateTime (", base::format(flows$ymd[1],
                                             format="%Z"), ")"), 
       title = paste(flows$model[1],"Flow Records, COMID:", flows$comid[1])) 
}
```

As part of a collaborative research effort, the multi-decadal reanalysis products from the [NOAA National Water Model](https://water.noaa.gov/about/nwm) (versions 1.2 and 2.0) have been reformatted and exposed through [CUAHSI](https://www.cuahsi.org/) at the [RENCI computing facility](https://renci.org/research/hydroshare/). This document outlines how someone can interact with the interface.

### TL;DR

`nwmHistoric` allows you to extract streamflow data from the NOAA National Water Model reanalysis products by NHD Catchment Identifier or USGS NWIS siteNumber:

```{r, eval = FALSE}
library(nwmHistoric)

# Extract NWM flow records for USGS-11120000
nwis_extract <- readNWMdata(siteNumber = 11120000)

# Extract NWM flow records for comid-101 and comid-17595383
nhd_extract <- readNWMdata(comid = c(101,17595383))
```

## Extraction

Reanalysis data can be found with the `readNWMdata` function. The API requires users to specify NHD Catchment Identifier(s) (COMID), or NWIS gage ID(s) for time series extraction.

### NHD Catchment Identifier (COMID)

NWM reanalysis data can be extracted by NHD Catchment Identifier(s). The following code pulls the complete reanalysis data archive for COMID 17595383, a NHD reach near Goleta, California.

```{r, eval = TRUE}
flows = readNWMdata(comid = 17595383)
head(flows)
dim(flows)
```

### USGS NWIS Site Number

Alternatively, data can be extracted using USGS NWIS site ID(s).

```{r, eval = TRUE}
flows = readNWMdata(siteNumber  = 11120000)
head(flows)
dim(flows)
```

## Contraint Parameters

### Date Range

By default, the API extracts the complete flow record for each ID. This range can constrained by providing a `startDate` and `endDate` in the form of ("YYYY-MM-DD"). 

```{r}
utc <- readNWMdata(comid = 17595383, 
                   startDate = "2010-01-01", 
                   endDate   = "2010-01-31")

# Check time dimensions
range(utc$time)
```

### Time-zone

NWM output is recorded in UTC. Instead, users can specify a timezone. For our COMID in Goleta, California we can request data in `US/Pacific` time. Valid times zone can be found with `OlsonNames()`.

```{r}
pst = readNWMdata(comid = 17595383, 
                  startDate = "2010-01-01", 
                  endDate = "2010-01-31", 
                  tz = "US/Pacific")

# Check time dimensions
range(pst$time)
```

Finally, we can plot the UTC return against the PST return:

```{r, echo = FALSE}
ggplot() +
  geom_line(data = utc, aes(x = time, y = flow, color = "UTC")) +
  geom_line(data = pst, aes(x = utc$time, y = flow, color = "PST")) +
  ggpubr::theme_pubr() +
  ggpubr::color_palette('aaas') +
  labs(x = 'Flows', y = "DateTime", title = "Time Zone Comparison", color = "Timezone") + theme(legend.position = "bottom")
```

### Model Version

The reanalysis product for versions 1.2 and 2.0 of the NWM are available. By default the API extracts data from version 2.0, but data from 1.2 can be requested as well.

```{r}
pst_v12 = readNWMdata(comid = 17595383, 
                      startDate = "2010-01-01", 
                      endDate   = "2010-01-31",
                      tz = "US/Pacific",
                      version =  1.2)
```

```{r, echo = FALSE}
ggplot() + 
  geom_line(data = pst_v12,   
            aes(x = time, y = flow, color = "v1.2")) + 
  geom_line(data = utc, 
            aes(x = time, y = flow, color = "v2.0")) +
  ggpubr::theme_pubr() +
  ggpubr::color_palette('aaas') +
  labs(x = 'Flows', y = "DateTime", title = "Model Comparison", 
       color = "Model") + 
  theme(legend.position = "bottom")
```
In section 2, we will look at how we can facilitate COMID and USGS NWIS discovery. In section 3 we will look at some techniques to handle time aggregations and splitting.


```{r teardown, include=FALSE}
options(oldoption)
```
