---
title: "Time Aggregation"
vignette: >
  %\VignetteIndexEntry{pkgdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup, echo = FALSE}
library(ggplot2)
library(dplyr)
library(viridis)
library(ggpubr)
library(nwmHistoric)
```

Raw reanalysis data comes in hourly time intervals. For many analyses, different time aggregates are needed (e.g. by year, by-month ...). To facilitate temporal aggregations and summaries, the `nwmHistoric` package provides a `split_time` function and a family of `aggregation` functions.


## Split-time

`split_time` in a convenience function to break the hourly time stamps into base temporal units. In addition to the basic year, month, day, and hour, the season, water year, and Day-of-Water Year (DOWY) are added: 

```{r}
# Get 5-years of data and split the time attribute
st = readNWMdata(comid = 101, 
                 startDate = "2005-01-01", 
                 endDate = "2009-12-31") %>%
  split_time()

head(st)
```
Using the split time attributes, we can plot the hourly flow records by water-year, grouped by water-year:

```{r}
ggplot(data = st, aes(x = DOWY, y = flow, color = flow)) + 
  geom_line(size = 1) +
  facet_grid(wy~.) +
  labs(y= "Daily Flow (cms)", x= "Day of Water Year", 
       title="Hourly Discharge for COMID 101") +
  scale_color_viridis_c() +
  theme_minimal()
```

## Aggregation

Often you might want to split and summarize your data, for example "average monthly flow" or "median annual flow". For these tasks, a family of aggregation methods allow users to define an temporal unit via the method name, and pass summarizing function(s) as parameters. Signiture names follow the pattern of `aggregate_*` where `*` represents the common date (and hydro-specific) symbols seen below.

| Symbol          | Aggregate          |
| --------------- |:------------------:|
| y               | year               |
| m               | month              |
| d               | day of moth        |
| doy             | day of year        |
| j               | Julian day         |
| s               | season             |
| wy              | water year         |
| dowy            | day of water year  |


These symbols can be combined to provide useful, common aggregation patterns, `r length(grep("aggregate",ls("package:nwmHistoric")))` of these are included the package (some are shown below):

| Aggregate Unit Symbol         |       Description    | 
| ----------------| :---------------------------------:| 
| *_record        | Entire Record                      | 
| *_y             | Year                               | 
| *_m             | Month                              | 
| *_j             | Julian Day                         | 
| *_s             | season                             | 
| *_wy            | Water Year                         | 
| *_ym            | Year and Month                     | 
| *_yj            | Year and Julian day                | 
| *_ymd           | Day of the Year                    | 
| *_ys            | Year and Season                    | 
| *_wym           | Water Year and Month               | 
| *_wymd          | Julian Day of the Water Year       | 
| *_wys           | Water Year and Season              |
| *_dowy          | Day of Water Year                  |

### Examples

First lets grab some data for a COMID found near Baton Rouge, LA along the Mississippi.

```{r}
flows = readNWMdata(comid = 19085515)
```

### Single function

Using the flow data grabbed above, we might be interested in seeing the monthly mean flow rates across the dataset. We can do this by passing the flow records to `aggregate_m` (m = month) and using `mean` as the function:

```{r}
# Aggregate hourly flows to monthly averages by year
monthly = aggregate_m(flows, mean)

ggplot(data = monthly) + 
  geom_col(aes(x = factor(month), y = flow)) +
  theme_pubr() + 
  labs(x = "Month", y = "Q (cms)", 
       title = 'Monthly Average 2005 - 2009')
```

Alternatively we might be interested in the monthly variability in each year. We can do this by passing the flow records to `aggregate_ym` (ym = year,month) and using `mean` as the summarizing function:

```{r}
# Aggregate hourly flows to monthly averages by year
ym = aggregate_ym(flows, mean)

ggplot(data = ym, aes(x = factor(month), y = flow)) + 
  geom_boxplot(alpha = .5) +
  theme_pubr() + 
  labs(x = "Month", y = "Q (cms)", 
       title = 'By month average 2005 - 2009')
```

```{r}
# Aggregate hourly flows to monthly averages by year
peak = aggregate_y(flows, 'max')

ggplot(data = peak, aes(x = year, y = flow)) + 
  geom_point(color = "navy", alpha = .5) +
  theme_pubr() + 
  labs(x = "Month", y = "Q (cms)", 
       title = 'Annual Peaks 2005 - 2009')
```

### Multiple functions

So far we have only looked at passing `mean` to `aggregate_*` but multiple functions can also be passed as a vector. The following code returns the seasonal (s) mean and standard deviation.

```{r}
# Aggregate by season
seasons = aggregate_s(flows, fun = c('mean', 'sd'))

ggplot(data = seasons, aes(x = season, y = flow_mean)) +
  geom_errorbar(aes(ymin=flow_mean - flow_sd, ymax=flow_mean + flow_sd, width = .25)) +
  geom_point(color = "darkred", size = 2) +
  theme_pubr() +
  labs(x  = "Season", y = "Q(cms)", 
       title = "Seasonal mean +- sd")
```


### Custom Functions

Equally important, you are not limited to base R functions. Instead you can pass any function to `fun` that works over a vector of streamflow elements. In the code below we ask for a number of percentiles along with some other summary statistics:

```{r}
# Aggregate by Julien Day
jul = aggregate_j(flows, fun = c(
                          n05 = function(x){quantile(x,.05)},
                          n25 = function(x){quantile(x,.25)},
                          n75 = function(x){quantile(x,.75)},
                          n95 = function(x){quantile(x,.95)},
                          median = median,
                          mean   = mean,
                          min    = min,
                          max    = max))


ggplot(data = jul, aes(x = julian)) +
  geom_ribbon(aes(ymin=flow_min, ymax=flow_max), fill="#B2DFEE") +
  geom_ribbon(aes(ymin=flow_n05, ymax=flow_n95), fill="#9AC0CD") +
  geom_ribbon(aes(ymin=flow_n25, ymax=flow_n75), fill="#68838B") +
  geom_line(aes(y = flow_mean), col = "#104E8B" ) +
  geom_line(aes(y = flow_median), col = "#AFEEEE") + 
  theme_pubr() +
  labs(x = "Day of Year", y = "Q (cms)", 
       title = "Flows Somewhere on the Mississippi...",
       subtitle = "26 years of record")
```

